### 2025-03-14 misc/expo/entropy.tex
Ben created

\proof The hypotheses give the three inequalities
$$\Eta\{X\} \le \Eta\{Z,W\}, \quad \Eta\{Y\}\le \Eta\{Z\},\quad\hbox{and}\quad \Eta\{Y\}\le \Eta\{W\}.$$
From this we see that
$$2\Eta\{X\} + 2\Eta\{Y\} \le 2\Eta\{Z,W\} + \Eta\{Z\} + \Eta\{W\}.$$
But then since conditioning does not increase entropy, we have
$$2\Eta\{X\} + 2\Eta\{Y\} \le 2\Eta\{Z\} + 2\Eta\{W\},$$
whence dividing both sides by $2$ completes the proof.\slug

The submodularity inequality is often stated in terms of a triple of random variables in terms of
the {\it conditional mutual information}, which is defined by
$$\eqalign{
\II\{ X : Y \given Z\} &= \sum_{z\in C} \pr\{Z = z\} \II\bigl\{ (X\given Z = z) : (Y\given Z = z)\bigr\} \cr
&= \Eta\{X\given Z\} - \Eta\{X\given Y, Z\} \cr
&= \Eta\{Y\given Z\} - \Eta\{Y\given X, Z\}.\cr
}$$

@@

By submodularity, we have $\II\bigl\{ (X-Y : Z) \given X-Z\bigr\} \ge 0$, so
\edef\eqfirstsubmodularity{\the\eqcount}
$$\eqalign{
0 &\le \II\bigl\{ (X-Y : Z) \given X-Z\bigr\} \cr
&\le \Eta\{X-Y \given X-Z\} + \Eta\{ Z\given X-Z\} - \Eta\{ X-Y, Z \given X-Z\} \cr
&\le \Eta\{X-Y, X-Z\} + \Eta\{ Z, X-Z\} - \Eta\{ X-Y, Z, X-Z\} - \Eta\{X-Z\}.\cr
}\adveq$$
Now, since the values $(x-y,x-z)$ taken by $(X-Y,X-Z)$ are in bijection with values
$(x-z,y-z)$ taken by $(X-Z, Y-Z)$ via the map $(v,w) \mapsto (w, w-v)$,
by the invariance axiom we have
$$\Eta\{X-Y, X-Z\} = \Eta\{X-Z, Y-Z\} \le \Eta\{X-Z\} + \Eta\{Y-Z\}.$$
Similar invocations of the invariance axiom give
$$\Eta\{Z, X-Z\} = \Eta\{X,Z\}$$
and
$$\Eta\{X-Y, Z, X-Z\} = \Eta\{X,Y,Z\} = \Eta\{X,Z\} + \Eta\{Y\},$$
where in the latter statement the second equality follows from the fact that $(X,Y)$ and $Z$ are
independent.
Substituting these three inequalities into~\refeq{\eqfirstsubmodularity}, we have
$$0\le \Eta\{X-Y\} + \Eta\{Y-Z\} +\Eta\{X,Z\} -\Eta\{X,Z\} + \Eta\{Y\} - \Eta\{X-Z\},$$
whence
$$ \Eta\{X-Z\} \le \Eta\{X-Y\} + \Eta\{Y-Z\} + \Eta\{Y\},$$
which completes the proof.\slug

@@

We prove the following lemma.

\proclaim Lemma \advthm.
Let $X$ and $Y$ be random variables taking values in $\FF_p^*$ for some prime $p$.
There exists $u\in \FF_p^*$ such that
$$\Eta\{X,uY\} \ge {\Eta\{X,Y\}\over 2^{\Etaseven\{X,Y\}}} - 1,$$

\proof Let $U$ be a random variable uniform on $\FF_p^*$ and independent of both $X$ and $Y$.
Independence gives us the identity $\Eta\{X+UY\} = \Eta\{U\} = \log(p-1)$.
We let $(X_1, Y_1)$, $(X_2,Y_2)$, and $U$ be conditionally independent trials
of $(X,Y)$, $(X,Y)$, and $U$ respectively, relative to $X+UY$.
Put
$$H = \Eta\{X_1, Y_1, X_2, Y_2, U\}.$$
Let $W = 1$ if $X_1 = X_2$ and $W=0$ otherwise. Then since $X_1 + UY_1 = X+UY$, we have
$$\eqalign{
H &= \Eta\{X_1, Y_1, X_2, Y_2, W, X+UY\} \cr
&= \Eta\{U\} + \Eta\{W\} + \Eta\{X_1, Y_1, X_2, Y_2 \given W, X+UY\}.\cr
}$$
If $X_1 = X_2$, then $Y_1 = Y_2$, so by conditional independence,
$$\eqalign{
H &= \Eta\{U\} + \Eta\{W\} +  \pr\{X_1 = X_2\given X+UY\} \Eta\{X_1, Y_1 \given X+UY\} \cr
&\qquad+ \pr\{X_1 \ne X_2 \given X+UY\} \Eta\{X_1,Y_1, X_2,Y_2\given X+UY\}.\cr
&= \Eta\{U\} + \Eta\{W\} + \pr\{X_1 = X_2\}\bigl( \Eta\{X,Y\given X+UY\}\bigr) \cr
&\qquad
+\pr\{X_1 \ne X_2\given X+UY\}\bigl( 2\Eta\{X,Y\given X+UY\} \bigr) \cr
&= \Eta\{U\} + \Eta\{W\} + \Eta\{X,Y\given X+UY\}\cr
&\qquad+\pr\{X_1 \ne X_2\given X+UY\}\Eta\{X,Y \given X+UY\} \cr
}$$
But
$$\Eta\{X,Y\given X+UY\} = \Eta\{X,Y, X+UY\} - \Eta\{X+UY\} = \Eta\{X,Y\}$$
by independence of $(X,Y)$ and $U$, so
$$H =\bigl(1+\pr\{X_1\ne X_2\given X+UY\}\}\bigr) \Eta\{X,Y\} + \Eta\{U\} + \Eta\{W\}.$$

On the other hand, by invertibility of each $u\in \FF_p^*$, we have
$$\eqalign{
H - \Eta\{U\} &= \Eta\{X_1, Y_1, X_2, Y_2\given U\} \cr
&= {1\over p-1} \sum_{u\in \FF_p^*} \Eta\{X_1, Y_1, X_2, Y_2\given U=u\} \cr
&= {1\over p-1} \sum_{u\in \FF_p^*} \Eta\{X_1, uY_1, X_2, uY_2\} \cr
&= {1\over p-1} \sum_{u\in \FF_p^*} \AA\{X, uY\}.\cr
}$$
Consequently there exists $u\in \FF_p^*$ such that
$$\eqalign{
\AA\{X,uY\} &\le H-\Eta\{U\} \cr
&= \bigl(1+\pr\{X_1\ne X_2\given X+UY\}\bigr) \Eta\{X,Y\} + \Eta\{W\} \cr
&\le \bigl(1+\pr\{X_1\ne X_2\given X+UY\}\bigr) \Eta\{X,Y\} + 1, \cr
}$$
and for this choice of $u$ we have
$$\Eta\{X+uY\} \ge \pr\{X_1 = X_2\given X+UY\} \Eta\{X,Y\} - 1.$$
But
$$\pr\{X_1 = X_2\given X+UY\} = \pr\bigl\{ (X_1,Y_1) = (X_2, Y_2) \given X+UY\bigr\},$$
and by [TODO: add lemma], we have
$$\eqalign{
\pr\{ (X_1,Y_1) = (X_2, Y_2) \given X+UY\} &\ge 2^{-\Etaseven\{X, Y \given X+UY\}} \cr
&= 2^{-\Etaseven\{X, Y\}},\cr
}$$
which immediately yields the desired inequality.\slug

>>>>

\proof Since $(Z,W)$ determines $X$, for each possible value $y$ that $Y$ takes,
$(Z,W\given Y= y)$ determines $(X\given Y=y)$. Summing over the support of $Y$ and
applying Corollary~{\cordetermines} gives us
$$ \Eta\{X\given Y\} \le \Eta\{Z,W \given Y\} \le \Eta\{Z\given Y\} + \Eta\{W\given Y\}.$$
Expanding the conditional entropies and rearranging yields
$$\Eta\{X,Y\} + \Eta\{Y\} \le \Eta\{Z,Y\} + \Eta\{W,Y\}.$$
But $Z$ and $W$ each determine $Y$, so $\Eta\{Z,Y\} = \Eta\{Z\}$ and
$\Eta\{W,Y\} = \Eta\{W\}$. Hence
$$\Eta\{X\} + \Eta\{Y\} \le \Eta\{X,Y\} + \Eta\{Y\} \le \Eta\{Z\} + \Eta\{W\},$$
as desired.\slug

@@

By submodularity, we have
$$\Eta\{X,Y,Z\} + \Eta\{X-Z\} \le \Eta\{X-Y, Y-Z\} + H\{Z, X-Z\}.$$
But $\Eta\{Z, X-Z\} = \Eta\{X,Z\}$, and
$\Eta\{X,Y,Z\} = \Eta\{X,Z\} + \Eta\{Y\}$ by independence, so
$$ \Eta\{X-Z\} + \Eta\{Y\} \le \Eta\{X-Y\} + \Eta\{Y-Z\}$$
and the result follows.\slug

@@

In this section, we prove an entropy variant of the Katz--Tao lemma~\ref{katztao}, which states
that if $R$ is a commutative ring, $R^*$ the set of elements in $R$ that are not zero divisors,
and $A\subseteq R^*$ has $|A+A|\le K|A|$ and $|A\cdot A|\le K|A|$, then there is a large subset
$A'$ of $A$ such that $|A'\cdot A' - A'\cdot A'| \le CK^C|A'|$, where $C$ is an absolute constant.

Crucial to the argumentation in this section is
the cancellative property of elements that are not zero divisors (if $xy = xz$ then $x(y-z) = 0$,
so for all $x\in R$ that is not a zero divisor, the function $y\mapsto xy$ is injective).
We require a special case of a rather technical lemma due to A.~M\'ath\'e and W.~L.~O'Regan.
We furnish a slightly simpler proof.

\proclaim Lemma \advthm. Let $X$, $Y$, $Z$, and $W$ be random variables taking values in $R^*$ for some
commutative ring $R$. Then
$$\eqalign{
\Eta\{XY + Z&W\} + \Eta\{X\} + \Eta\{Y\} + 2\Eta\{Z\} + 2\Eta\{W\} \cr
&\le \Eta\{X+Y\} + \Eta\{Z-W\} + \Eta\{XZ\} + 2\Eta\{YZ\} + 2\Eta\{ZW\}.\cr
}$$

\proof Let $S = \bigl( XY, ZW, (X+Z)Y, Z(Y-W)\bigr)$ for short. The tuples $(S,Y)$ and
$(X,Y,Z,W)$ determine each other, so $\Eta\{S,Y\} = \Eta\{X,Y,Z,W\}$. By submodularity,
$$\Eta\{S,Y\} + \Eta\{XY\} \le \Eta\{S\} + \Eta\{X,Y\},$$
so
\edef\eqfirstsub{\the\eqcount}
$$\eqalign{
\Eta\{S\} &\ge \Eta\{X,Y,Z,W\} + \Eta\{XY\} - \Eta\{X,Y\} \cr
&= \Eta\{Z,W\} + \Eta\{XY\}, \cr
}\adveq$$
by independence.
Another application of submodularity yields
$$\Eta\{XY+ZW\} + \Eta\{S\} \le \Eta\{XY, ZW\} + \Eta\bigl\{ (X+Z)Y, Z(Y-W)\bigr\},$$
since $(X+Z)Y + Z(Y-W) = XY+ZW$. Substituting in~\refeq{\eqfirstsub}, we get
\edef\eqsecondsub{\the\eqcount}
$$\Eta\{XY+ZW\} + \Eta\{Z,W\} \le \Eta\{ZW\} + \Eta\bigl\{ (X+Z)Y\bigr\} + \Eta\bigl\{Z(Y-W)\bigr\},\adveq$$
where we have cancelled a $\Eta\{XY\}$ term from both sides.
Lastly, note that
$$\eqalign{
\Eta\bigl\{ (X+Z)Y\bigr\} &\le \Eta\{XZ, YZ\} + \Eta\{X+Z,Y\} - \Eta\{X,Y,Z\} \cr
&\le \Eta\{XZ\} + \Eta\{YZ\} + \Eta\{X+Z\} - \Eta\{X\} - \Eta\{Z\} \cr
}$$
by submodularity, and likewise
$$\eqalign{
\Eta\bigl\{ Z(Y-W)\bigr\} &\le \Eta\{YZ, ZW\} + \Eta\{Z,Y-W\} - \Eta\{Y,Z,W\} \cr
&\le \Eta\{YZ\} + \Eta\{ZW\} + \Eta\{Y-W\} - \Eta\{Y\} - \Eta\{W\}.\cr
}$$
Substituting both of these inequalities into~\refeq{\eqsecondsub}, gives exactly
the desired inequality.\slug

\parenproclaim Lemma {\advthm} (Entropic Katz--Tao lemma).
Let $X$ be a random variable taking values in $F^*$ for some field $F$, and let
$X_1$, $X_2$, $X_3$, and $X_4$ be independent copies of $X$. If
$$\max\bigl(\Eta\{X_1+X_2\}, \Eta\{X_1X_2\}\bigr) \le \Eta\{X\} + \log K .$$
for some constant $K$, then
$$\Eta\{X_1 X_2 + X_3X_4\} \le \Eta\{X\} + 8\log K.\adveq$$

\proof The previous lemma with $X\gets X_1$, $Y\gets X_2$, $Z\gets X_3$ and $W\gets X_4$
gives
$$\Eta\{X_1X_2+X_3X_4\} \le \Eta\{X_1+X_2\} + \Eta\{X_1-X_2\} + 5\Eta\{X_1X_2\} - 6\Eta\{X\},$$
after applying the fact that the $X_i$ are identically distributed.
By the Ruzsa triangle inequality, we have
$$\Eta\{X_1 - X_2\} \le \Eta\{X_1 + X_3\} + \Eta\{X_2 + X_3\} - \Eta\{X_3\},$$
so
$$\Eta\{X_1X_2+X_3X_4\} \le 3\Eta\{X_1+X_2\} + 5\Eta\{X_1X_2\} - 7\Eta\{X\}.$$
Finally, we use the hypothesis
$\max\bigl(\Eta\{X_1+X_2\}, \Eta\{X_1X_2\}\bigr) \le \Eta\{X\} + \log K$ to arrive
at the required bound.\slug
